\section{Validity policies}

In a previous section of this chapter, we mentionned that $\isvalid{\phi}{.}$ 
is a function if for every integer variable $i$, resp. pointer variable
$x$, and at each abstract state of the program there exists one and only one evaluation for $\isvalid{\phi}{i}$, resp. $\isvalid{\phi}{x}$.

Since the evaluation of the validity of each variable $i$, $\isvalid{\phi}{i}$, 
on any next state is deduced from the current state, the evaluation of
the validity of a variable only depends of the initial value of $\isvalid{\phi}{i}$
at the entry point of the considered function.

We introduce a policy for initializing the validity of each integer and each
pointer variable, and we defend why we fix such values.
We consider tree subset of variables :

\begin{itemize}
\item The set of global variables,
\item the set of parameters, also called "formals" in Cil,
\item the set of local variabes, stored on the stack.
\end{itemize}

Here is the initial value assigned to, w.l.o.g. $\isvalid{\Formula{\true}{\Emp}}{i}$ at the entry point of the function (TODO : Explique plus clairement ce
aue l'on entend par "point d'entree de la fonction etudiee"):

$$
\begin{array}{|l|l|}
\hline
\mbox{Location} & \mbox{Initial evaluation} \\
\hline
\mbox{Global}& \dk \\
\mbox{Parameters} & \dk \\
\mbox{Locals} & \false \\
\hline
\end{array}
$$


\subsection{propagation}

\textbf{Assignation}

\textbf{Assignation of the returned value of a function call}

\textbf{Global variables : Monothread case, multithread case.}
\begin{itemize}
\item Monothread case : If a global variable is assigned a value
from a valid variable or expression that have the same type, then
its validity is set to true. However, If the control reaches a 
function call, then this validity shall be reseted to $\dk$, unless
one can garantee that the called function does not perform an access
to this variable -- which goes beyond the scope of the current analysis. 

\item Multithread case : Even if a global variable is assigned a valid
value, it can be modified as soon as the control flow is peempted to another
thread, that might perform an access to the global variable. In this case,
the validity of this  function shall be set as the constant $\dk$.
\end{itemize}
